---
title: "Survival Analysis"
output: html_document
date: "2023-07-22"
author: "Mercedeh Movassagh PhD, Yale University"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview of this R Markdown
This R Markdown will teach you how to do survival analysis using R.

```{r  message=FALSE, warning=FALSE}
library(ggpubr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(MASS)
library(ISLR2)
library(car)
library(survival)
```

## Survival Analysis

For this analysis we will use the BrainCancer data in the ISRL2 package.

As a reminder status=1 indicates uncensored observation and status =0 indicates a censored observation. In this data set 35 patients die before the end of the study.The survival analysis is commonly done using the survifit().

```{r  message=FALSE, warning=FALSE}
#get familiar with the dataset
attach(BrainCancer)
names(BrainCancer)
table(sex)
table(status)
#Note time corresponds to yi (the time to the ith event, either censoring or death)
fit_surv<-survfit(Surv(time,status) ~ 1)
#Now olor the fit_surv
plot(fit_surv,xlab="Months",ylab="Estimated Probability of Survival")
#Now you can stratify by other features such as sex
fit_surv<-survfit(Surv(time,status) ~ sex)
plot(fit_surv,xlab="Months",ylab="Estimated Probability of Survival", col=c(2,4))
```

## Log-rank test

You can use the long rank test to compare survival between two groups in this example sex of the patients.

```{r  message=FALSE, warning=FALSE}
logrank_test<-survdiff(Surv(time,status) ~ sex)
logrank_test
logrank_test$chisq
#The test shows there is no evidence of difference for survival between males and females.
```

## Cox Proprtional Hazardz

Recall cox-ph models can be used for both binary/categorical and continuous predictors for estimating survival.

```{r  message=FALSE, warning=FALSE}
fit_cox<-coxph(Surv(time,status) ~ sex)
summary(fit_cox)#Neither of the tests show association between sex and survival
#Now check for all other predictors
fit_all<-coxph(Surv(time, status) ~ sex + diagnosis + loc + ki + gtv + stereo)
fit_all
#Show all the predictors in one plot
all_new<-data.frame(
  diagnosis=levels(diagnosis),
  sex= rep("Female",4),
  loc= rep("Supratentorial",4),
  ki= rep(mean(ki),4),
  gtv = rep(mean(gtv),4),
  stereo = rep ("SRT",4)
)
Survplots<-survfit(fit_all,newdata = all_new)      
plot(Survplots, xlab = "Months",
     ylab= "Survival Probability", col= 2:5)
legend("bottomleft",levels(diagnosis), col=2:5, lty=1)
```
## Reference

reference: "James, G., D. Witten, T. Hastie, and R. Tibshirani. 2013. “An Introduction to Statistical Learning.” https://link.springer.com/content/pdf/10.1007/978-1-0716-1418-1.pdf."


